---
title: "Gibbs Sampling with Data Augmentation for Bayesian Analysis of Binary and Polychotomous Response Data"
subtitle: "STAT 230 Final Report"
author: "Damon Bayer & Corey Katz"
date: "12/16/2019"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: kable
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
source(here("plotting_functions.R"))

post_sample_summary <- function(post) {
  post %>% 
  pivot_longer(everything()) %>% 
  rename(Variable = name) %>% 
  group_by(Variable) %>% 
  summarize_all(list(mean = mean,
                     sd = sd,
                     "2.5%" = function(x) quantile(x, probs = 0.025),
                     "50%" = median,
                     "97.5%" = function(x) quantile(x, probs = 0.975)))
}
```

# Abstract

Here's the abstract.

# Introduction
<!-- (explaining the motivation behind the new algorithm) -->

We cite @albertchib93

Here's the introduction.

# Methodology
<!-- (explaining the algorithm), -->

Here's the methodology.

# Results
<!-- (applying your algorithm to simulated or real data) -->

For three datasets, we apply the appropriate previously discussed model as well as standard maximum likelihood estimation and custom models written in Stan [@mcstan] and impelemented in the RStan package [@rstan].

Here's the results.

## Small-cell Carcinoma

```{r Ordered Multinomial,echo=FALSE}
ordered_multinomial_post <- read_rds(here("ordered_multinomial.rds"))

post_sample_summary(ordered_multinomial_post)

trace_plot(ordered_multinomial_post)
dist_plot(ordered_multinomial_post)
```


# Discussion
<!-- (shortcomings of this new algorithm and how it can be improved). -->

Here's the discussion.

\pagebreak

# References

::: {#refs}
:::

\pagebreak

# Appendix